Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering

### 2017-06-18 R/workflow_oldVersion.Rmd
clean up and add biocworkflow dir

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
#bioconductor:
library(scRNAseq)
library(scone)
library(clusterExperiment)
library(BiocParallel)
#github
library(zinbwave)
library(slingshot)
#CRAN
library(doParallel)
library(ggplot2)
library(gplots)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(RColorBrewer)
library(digest)
library(rARPACK)
set.seed(20)
if(packageVersion("clusterExperiment")<'1.3.0.9009') stop("must have current develop version to avoid bugs")
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

@@

```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
knitr::include_graphics('../vignettes/workflow.png')
```

>>>>

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
#bioconductor:
library(scRNAseq)
library(scone)
library(clusterExperiment)
library(BiocParallel)
#github
library(zinbwave)
library(slingshot)
#CRAN
library(doParallel)
library(ggplot2)
library(gplots)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(RColorBrewer)
library(digest)
library(rARPACK)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

FP: this is the old version of the workflow, please use file workflow/workflow.Rmd that I created from package BiocWorkflowTools. Directory workflow in now all compatible with the bioc workflow format and we can directly convert the Rmd file to a tex file that can then be exported to overleaf for direct submission to F1000Research.

@@

```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
knitr::include_graphics('../workflow/workflow.png')
```

### 2017-06-18 workflow/workflow.Rmd
clean up and add biocworkflow dir

---
title: "Worflow"
author: "Workflow team"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    code_folding: hide
    toc_float: yes
---

@@

```{r}
runZinb <- TRUE
runClus <- TRUE
NCORES <- 7
```

FP: this is the old version of the workflow, please use file workflow/workflow.Rmd that I created from package BiocWorkflowTools. Directory workflow in now all compatible with the bioc workflow format and we can directly convert the Rmd file to a tex file that can then be exported to overleaf for direct submission to F1000Research.

FP: Version of zinbwave to use is last version of branch normvalues (branched from master). Parallel computing is now handle by BiocParallel. If you have a Windows machine, please update the code to allow paralell computing.

```{r parallel}
mysystem = Sys.info()[['sysname']]
switch(mysystem,
       Windows = {print("I'm a Windows PC and I'll use the snow for parallel computing (warning: it might be slower than serial computations).")},
       Linux = {print("I'm a penguin and I'll use package MulticoreParam for parallel computing.")},
       Darwin = {print("I'm a Mac and I'll use package doParallel for parallel computing.")})

if (mysystem == 'Darwin'){
  registerDoParallel(NCORES)
  register(DoparParam())
}else if (mysystem == 'Linux'){
  register(bpstart(MulticoreParam(workers=NCORES)))
}else{
  print('Please change this to allow parallel computing')
  register(SerialParam())
}
```

@@

# Steps of the workflow

We propose a worklow to analyze single cell RNA-Seq with the following steps

- Dimensionality reduction using zinbwave to get W which should capture the biology,
- Cluster cells using clusterExperiment on W to get the cluster labels,   
- Get lineage using slingshot on W and cluster labels from clusterExperiment,  
- Get DE genes between lineages/clusters.

Along the worflow, use deviance residuals as adjusted values.

```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Worflow to analyze single cell RNASeq data'}
knitr::include_graphics('../workflow/workflow.png')
```

## 1. Create a SummarizedExperiment object

Along the workflow, we want to use a unique SummarizedExperiement object carrying all the data we need.

EAP: I have updated the code to pull from a dataset on the repos that is created with the `createData.R` file. For now, I am filtering to the top 1000 most variable genes there, though we might want to add that to the code for the article. This will be slightly different data from Russell's, which didn't use all of the samples. We can adjust that decision later, or just compare the samples that are the same. (Russell's clusterLabels are in the meta data)

EAP: zinbFit doesn't accept data.frame objects, so currently have to have a `data.matrix` command. Should it be changed so that it does?

DR: I have changed the `createData.R` file so that it uses scone to filter out low-quality samples.

@@

load("../data/oe_se_1000Var.rda")
```

Here we only look at the 1000 most variable genes. EAP: see note above, I've commented out the filtering and added it to the `createData.R` for now.

@@

Cells have been processed in `r length(unique(batch))` different batches

@@

We have qc measures from the data

@@

In original work (FP: add ref), cells have been clustered into `r length(unique(clus.labels))` different clusters

@@

Batches are kind of confounded with the biology

@@

We have `r ncol(core)` cells.

```{r lookatdata}
dim(core)
assay(core)[1:3, 1:3]
```

## 2. Dimensionality reduction adjusting for gene and cell-level covariates

To cluster and get lineages we want to reduce the dimension of the data. We are going to use zinbwave to do so. First, let's fit zinbwave with first K = 0 to compute normalized values (i.e. deviance residuals) adjusted for batches. We could also adjust for gene length or GC content here. We then fit zinbwave to get the dimensionality reduced matrix W specifying the number of dimension K = 50. Eventually, we will call zinbwave just once where we would have an argument in zinbFit like "compute_normalized_values" in c(TRUE, FALSE). For K = 0 and K = 50, we correct for batch effect including batches in X.

@@

We use deviance residuals as normalized values for visualization. FP: explain rational: K=0 so residuals capture the bio adjusting for batch. Let's check that deviance residuals look ok.

@@

Boxplot of the normalized values for each cell. It seems that correction for batches is ok.

@@

PCA on the normalized values where color are for batches on the left and previously found clusters on the right. We want no clustering on the left side and clustering on the right side.

@@

@@

Let's check that performing MDS on W we have something coherent with original clusters.

@@

## 2. Clustering of the cells

@@

## 4. Pseudotime ordering

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC / immature neurons / MV 1 | green | original 2, 3, 11
c3 | Sus | red | original 4, 7
c4 | Neuron | orange | original 9, 12
c5 | Immature Neuron | purple | original 10, 14
c6 | new and small | brown | original 9
c7 | Microvillus | cyan | original 15

@@

## 5. DE analysis

@@

FP: Kelly, is it you who did the DE analysis for Russell paper? If yes, what tool did you use? On what data? The full quantile normalized counts? Do you have code available?

# Session Info

@@

>>>>

---
title: "Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering"
author: 
  - name: Author Name1
    affiliation: Address of author1
  - name: Author Name2
    affiliation: Address of author2
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
bibliography: ref.bib
output: BiocWorkflowTools::f1000_article
---

@@

```{r parallel}
runZinb <- TRUE
runClus <- TRUE
NCORES <- 7
mysystem = Sys.info()[['sysname']]
if (mysystem == 'Darwin'){
  registerDoParallel(NCORES)
  register(DoparParam())
}else if (mysystem == 'Linux'){
  register(bpstart(MulticoreParam(workers=NCORES)))
}else{
  print('Please change this to allow parallel computing')
  register(SerialParam())
}
```

@@

# Introduction

- growing interest in single-cell RNA-seq data. New statistical tools need to be developped compared to bulk RNA-seq because bias: drop-outs, amplification bias. There is also increasing size of datasets (10X-genomics, 1.3M cells) -> need of integrated workflows of new statistical tools
- previously developped : F1000 Research bioc [@Lun2016] and scater [@McCarthy2017] are pipelines containing several useful methods for quality control, visualisation and pre-processing of data. In these pipelines, single-cell expression data are organized in objects of the SCESet class which allows integrated workflows. However, they are used mostly to prepare the data for further downstream analysis and do not focus on steps like clustering, pseudotime ordering, and DE analysis.
- we propose an integrated workflow with dowstream analysis: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resumpling, pseudotime ordering, and DE analysis for the clusters found. We use a summarizedExperiment object to have an integrated pipeline. We also propose useful functions for visualization.

```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Our workflow'}
knitr::include_graphics('schema_workflow.png')
```

# Analysis of olfactory stem cell lineages using single cell RNA-seq data
## Overview

The workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs) which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.

The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium. To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, different lineages were assigned to each cell in the dataset and results were confirmed experimentally using in vivo lineage tracing. It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either sustentacular (support) cells or GBCs. Then, the GBC lineage, in turn, branches to give rise to olfactory sensory neurons, microvillous cells, and cells of the Bowman’s gland.

```{r stemcelldiff, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. Figure 2e from Fletcher et al paper.'}
knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
```

## Generation of the data

The dataset we use for the workflow is publicly available at https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz. It contains 849 cells and 28361 detected genes. Each cell has been labelled with labels publicly available at https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt. Pre-processing steps are as follow:
- ERCC and CreER cells were removed,
- low-quality cells were removed using package scone (ADD REF) and

FP: we should add details here.

Finally to speed up the computations, we kept only the 1000 most variable genes in this analysis. After the preprocessing steps, the dataset we are working with has 1000 genes and 747 cells. Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object

@@

load("../data/oe_se_1000Var.rda")
core
```

@@

Metadata for the cells is stored in colData from the SummarizedExperiment object. Cells have been processed in `r length(unique(batch))` different batches

@@

We also have access to the quality control measures that were used to filter low quality cells

@@

In original work [@Fletcher2017], cells have been clustered into `r length(unique(clus.labels))` different clusters

@@

Note that in this dataset batches are somehow confounded with the clusters found in the original work.

@@

## Dimensionality reduction adjusting for batches

In single-cell RNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed. It allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest.

Here, we perform dimensionality reduction using bioconductor package zinbwave which allows to fit a zero-inflated negative binomial model (ZINB-WaVE) to get a low-dimensional representation of the data while accounting for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.

As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and we adjust for the batches.

FP: do we want to expose the user to the equations? In the one hand, it would be easier to explain what we are looking at (W, normalized values, ...). In the other hand, it would need a good amount of explanations. I have a preference to show the equations, but what do you think?

@@

When function zinbDimRed is called, normalized values of the counts can be computed and used for visualization (e.g. in heatmaps). The normalized values are the residuals of the fit of our ZINB model where the gene and cell covariates were the same as the user specified in the zinbDimRed call but the number of unknown covariates is set to be null (i.e. K=0). When no unknown covariates are included in the model, the residuals represents the counts adjusted for the gene and cell covariates. To compute the residuals, we use the deviance residuals.

@@

As expected, the distributions of the normalized counts are independent of the batches

@@

We performed principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right. As expected, there are no clustering on the left side and clustering on the right side.

@@

Overall, looking at the normalized values, it seems that the batch effect was removed.

@@

Calling function zinbDimRed, we performed dimensionality reduction with K = 50. We can visualize the low dimensional matrix performing multidimensional scaling (MDS) in two dimensions. To make sure that the low dimensional matrix captured the signal of interest, the colors shown here are from the original clusters.

@@

## Clustering of the cells

@@

## Pseudotime ordering

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | immature neurons | MV 1 | green | original 2, 3, 11
c3 | Sus | red | original 4, 7
c4 | Neuron | orange | original 9, 12
c5 | Immature Neuron | purple | original 10, 14
c6 | new and small | brown | original 9
c7 | Microvillus | cyan | original 15

@@

## DE analysis

@@

## Further developments

- DE analysis using slingshot and zinbwave
- SingleCellExperiment class

# Conclusions

This workflow provides guidance to perform downstream analysis of single-cell RNA-seq datasets in R. We propose a worflow in four steps: dimensionality reduction while adjusting for gene and cell-level covariates, robust and stable clustering, pseudotime ordering, and DE analysis between the clusters. The workflow is general and flexible allowing the user to sustitute each step of the worflow with a different R tool and to input data that have been previously processed by different tools. We hope the workflow will ease the technical aspect of single-cell RNA-seq data analysis to help discovering new biological insights.

# Software availability
This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.

1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
5. Software license (AUTHOR TO PROVIDE; required)

The three packages zinbwave, clusterExperiment, and slingshot used in the workflow are bioconductor packages and are available on github at respectively https://github.com/drisso/zinbwave, https://github.com/epurdom/clusterExperiment, and https://github.com/kstreet13/slingshot. The source code for this package can be found at https://github.com/fperraudeau/singlecellworkflow under license XXX.

@@

# Author contributions
In order to give appropriate credit to each author of an article, the individual contributions of each author to the manuscript should be detailed in this section. We recommend using author initials and then stating briefly how they contributed.

# Competing interests
All financial, personal, or professional competing interests for any of the authors that could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article. If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.

# Grant information
Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc. Please do not list funding that you have that is not relevant to this specific piece of research. For each funder, please state the funder’s name, the grant number where applicable, and the individual to whom the grant was assigned. If your work was not funded by any grants, please include the line: 'The author(s) declared that no grants were involved in supporting this work.'

# Acknowledgments
This section should acknowledge anyone who contributed to the research or the
article but who does not qualify as an author based on the criteria provided earlier (e.g. someone or an organization that provided writing assistance). Please state how they contributed; authors should obtain permission to acknowledge from all those mentioned in the Acknowledgments section.
